---
title: "Verify the Claim: Before the world changed: New census data looks at pre-Covid San Francisco"
author: "Sara Cerami"
date: September 18, 2026
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
  execute:
    warning: false
    message: false
---
    
In 2022, the SF Standard, a San Francisco news outlet, published an article entitled, "Before the world changed: New census data looks at pre-Covid San Francisco: The latest ACS five-year study gives robust information on the city’s neighborhoods." It was comprised of a number of various maps and accompanying reports documenting how San Francisco's demographics shifted between the recently released 5-year American Community Survey (2016-2020) and the prior most recent non overlapping survey (2011-2015). The first map highlights the percent change of average household income in each San Francisco zipcode between the two 5-year ACS reports. It does not mention its methodology or approach to analyzing the data. I am interested in knowing if it was incorporated or addressed in the calculation of the percent change.        

The article does not explicitly say what variable it is using to calculate the average household income per zipcode, so to be sure I was on the right track, I pulled both the median household income as well as the mean household income (DP03_0063E). Comparing the two forms of averages, it was evident that the article used mean. The mean household income data accounts for inflation, calculated based on the year the report is issued. Additionally, median household income is top-coded in 5-year ACS tract level data, meaning that there is a cap on the highest income to protect privacy. While it may not impact the true median, it is a notable decision from the Census Bureau.  

The two ACS reports do not overlap.

```{r}
#| include: false
library(tidycensus)
library(tidyverse)

sf_tract_income <- get_acs(
  geography = "tract",
  variables = "DP03_0063E",
  state = "CA",
  county = "San Francisco",
  year = 2020,
  survey = "acs5"
)
```

```{r}
#| include: false
sf_tract_income <- sf_tract_income %>%
  mutate(moe_pct = moe / estimate * 100)

sf_tract_income <- sf_tract_income %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 10 ~ "Moderate",
    TRUE         ~ "Low confidence"
    ))
```

```{r}
#| eval: false
ggplot(sf_tract_income)+
  aes(x=estimate, y = moe_pct)+
  geom_point()
```

```{r}
#| eval: false
ggplot(sf_tract_income)+
  aes(x=estimate, y = moe_pct)+
  geom_point(color = "maroon")
```

```{r}
#| include: false
sf_tract_income %>%
  ggplot(aes(x = NAME, y = estimate)) +
  geom_col() +
  geom_errorbar(aes(ymin = estimate - moe, ymax = estimate + moe))
  
sf_tract_income %>%
  arrange(desc(moe_pct)) %>%
  mutate(NAME = str_remove(NAME, "San Francisco County, California")) %>%
  mutate(NAME = str_remove(NAME, "Census")) %>%
  slice_head(n = 15) %>%
  ggplot(aes(x = reorder(NAME, moe), y = estimate)) +
  geom_col(fill = "maroon") +
  geom_errorbar(aes(ymin = estimate - moe, ymax = estimate + moe), width = 0.3) +
  coord_flip() +
  theme_minimal() +
  theme(
    plot.title = element_text(
      family = "sans",
      face = "bold",
      size = 12,
      color = "black",
      hjust = 0,
      vjust = 0,
      )
  ) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "SF Mean Household Income Estimates",
    subtitle = "San Francisco Census Tracts, ACS 2020 5-year estimates",
    x = "Tract",
    y = "Mean Household Income and MOE"
    )
```

Because I am not calculating a rate from two separate counts, and because mean is an aggregate of data, not a count, moe_prop() does not apply here. 

The NA data is located in tracts that are primarily parkland, pacific ocean, or bay. I decided to keep them in, because they all have some level of data in one of the ACS reports.   

```{r}
#| include: false
get_mean_hh_income_sf <- function(yr) {
  get_acs(
    geography = "tract",
    variables = "DP03_0063E",
    county = "San Francisco",
    state     = "CA",
    year      = yr,
    survey    = "acs5"
  ) %>%
    mutate(period = paste0(yr - 4, "-", yr))
}

mean_hh_income_sf_change <- bind_rows(
  get_mean_hh_income_sf(2015),
  get_mean_hh_income_sf(2020)
)

mean_hh_income_sf_new <- mean_hh_income_sf_change %>%
  select(GEOID, NAME, period, estimate, moe) %>%
  pivot_wider(
    names_from  = period,
    values_from = c(estimate, moe)
  )

hh_income_tested <- mean_hh_income_sf_new %>%
  mutate(
    change   = `estimate_2016-2020` - `estimate_2011-2015`,
    se_1     = `moe_2011-2015` / 1.645,
    se_2     = `moe_2016-2020` / 1.645,
    se_diff  = sqrt(se_1^2 + se_2^2),
    moe_diff = se_diff * 1.645,
    z        = change / se_diff,
    significant = abs(z) > 1.645
  )
  
hh_income_tested %>%
  count(significant)
```

```{r}
#| include: false
library(gt)
```

```{r}
#| include: false
reliability_table <- sf_tract_income %>%
  mutate(
    cv = (moe / 1.645) / estimate * 100,
    dot = case_when(
      cv < 12  ~ "🟢",
      cv <= 40 ~ "🟡",
      TRUE     ~ "🔴"
    ),
    flag = case_when(
      cv < 12  ~ "Reliable",
      cv <= 40 ~ "Somewhat reliable",
      TRUE     ~ "Unreliable"
    ),
    NAME = str_remove(NAME, " San Francisco County, California")
  ) %>%
  arrange(desc(cv)) %>%
  select(dot, NAME, estimate, moe, cv, flag)

reliability_table %>%
  gt() %>%
  fmt_number(columns = c(estimate, moe), decimals = 0) %>%
  fmt_number(columns = cv, decimals = 1) %>%
  cols_label(
    dot      = "",
    NAME     = "Tract",
    estimate = "Estimate",
    moe      = "MOE",
    cv       = "CV (%)",
    flag     = "Reliability"
  ) %>%
  cols_align(align = "center", columns = dot) %>%
  tab_header(
    title    = "Mean Household Income by San Francisco Census Tract",
    subtitle = "San Francisco Census Tracts, ACS 2020 5-year estimates"
  ) %>%
  tab_source_note(
    "Reliability thresholds follow Jurjevich et al. (2018): CV < 12% reliable, 12–40% somewhat reliable, > 40% unreliable."
  )
```

```{r}
#| eval: false
acs_2015 <- get_acs(
  geography = "tract",
  county = "San Francisco",
  variables = "DP03_0063E",
  year = 2015,
  state = "CA",
  geometry = TRUE,
)

acs_2020 <- get_acs(
  geography = "tract",
  county = "San Francisco",
  variables = "DP03_0063E",
  year = 2020,
  state = "CA",
  geometry = FALSE
)

library(dplyr)
library(sf)
```

```{r}
#| eval: false
hh_income_tested <- hh_income_tested %>%
  left_join(acs_2015 %>% select(GEOID, geometry), by = 'GEOID')
```

```{r}
#| include: false
hh_income_tested <- hh_income_tested %>%
  filter(NAME != "Census Tract 9804.01, San Francisco County, California", NAME != "Census Tract 017903, San Francisco County, California")

library(ggplot2)
geom_to_sf <- st_as_sf(hh_income_tested)
```

```{r}
ggplot(data = geom_to_sf) +
  geom_sf(aes(fill = significant), color = "black", linewidth = .2) +
  labs(title = "Tract by Statistical Significance", subtitle = "Assessing the Statistical Significance of Mean Household \nIncome Changes: ACS 2015 vs. ACS 2020",  fill = "Legend") +
  scale_fill_manual(values = c("FALSE" = "limegreen", "TRUE" = "maroon")) +
  theme_void() +
  theme(
    plot.title = element_text(
      family = "sans",
      face = "bold",
      size = 16,
      color = "black",
      hjust = 0,
      vjust = 0,
      margin = margin(b=8)
    ),
    plot.subtitle = element_text(
      family = "sans",
      face = "italic",
      hjust = 0,
      vjust = 0,
      margin = margin(b=-90)
    )
  )
```

The article's broadest claim is that each Zip code in San Francisco saw an increase in mean household income between the ACS15 and ACS20. Based on my analysis, it does not seem as if the article took the margin of error into account when providing estimates of rate of income increase. Of the 155 tracts with estimates in both periods, 117 show an increase in mean household income that is statistically distinguishable from zero at 90% confidence, while 38 show changes that fall within their margins of error.

Several claims cannot be determined from this analysis. The specific percentage gains, over 75% in 94158 and 94102 and over 50% in 94118, 94131, and 94123, are reported for ZIP codes, while my analysis used tracts.
